Follow-up Samples

### 2017-01-13 submission/manuscript_outline_20161024.tex
add pdf with completed results

\vspace{25mm}

\begin{center}
Running Title: Human Microbiome and Colorectal Cancer

@@

Marc A Sze, Nielson T Baxter, Mack T Ruffin IV, Mary AM Rogers, and Patrick D Schloss${^\dagger}$

Contributions: All authors were involved in the conception and design of the study. MAS analyzed the data. NTB processed samples and analyzed the data. All authors interpreted the data. MAS and PDS wrote the manuscript. All authors reviewed and revised the manuscript. All authors read and approved the final manuscript.

@@

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI
\end{center}

@@

Colorectal cancer (CRC) continues to be a worldwide health problem with
early detection being used as a key component in mitigating deaths due
to the disease. Previous research suggests a link between stool
bacterial microbiome and CRC. In this study, we used a model based on
the microbiome, demographics, and prior medical history to classify
individuals as having a lesion (i.e.~adenoma or carcinoma). We then used
this model to characterize the change in the gut microbiota before and
after surgery. The overall objective was to investigate the changes in
the microbiome after surgery in patients with adenomas or carcinomas.
This model was tested on a 66 person group that included samples before
and after treatment to allow for the assessment of how the model adjusts
risk after treatment. The model used for prediction had an AUC of NA.
For the follow up samples our Random Forest model significantly
decreased the positive probability of lesion compared to the initial
samples for both adenoma (P-value = 3.64e-11) and carcinoma (P-value =
7.95e-08). Our model predicted that 36.4\% of the 67-person cohort had
normal colons and 63.6\% had a lesion. Some OTUs that changed the most
before and after treatment included OTUs that were affiliated with
members of Blautia, Streptococcus and Escherichia/Shigella. Our model
suggests that treatment does significantly reduce the probability of
having a colonic lesion. Further surveillance of these individuals will
enable us to determine whether models such as the one we present here
can also be used to predict recurrence of colorectal cancer.

@@

\subsubsection{Importance}\label{importance}

This is one of the first studies to investigate within humans what
happens to the bacterial microbiome before and after adenoma and
carcinoma treatment. Specifically, it aims to assess how a random forest
machine learning algorithm built model respond to treatment and adjust
it's positive probability calls of whether the individual has an adenoma
or carcinoma due to surgical removal of the lesion.

@@

\subsection{Introduction}\label{introduction}

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age (1, 2). Over the last few years death due to the
disease has seen a significant decrease thanks mainly to improvements in
screening (1). However, despite this giant improvement there are still
approximately 50,000 deaths from the disease a year (2). It is estimated
that around 5-10\% of all CRCs can be explained by autosomal dominant
inheritance (3). The vast majority of CRCs are not inherited and the
exact etiology to disease has not been well worked out (2). Although
many risk factors have been identified (2) and non-invasive screening
techniques have started to be put into consistent use (4, 5) there has
been a consistent increase in the incidence of CRC in the younger
population.

This increased incidence of CRC in the younger population is concerncing
since having either an adenoma or carcinoma increases ones risk for
future adenomas or carcinomas (6--8). This increased risk can also carry
with it an increased risk of mortality due to this recurrence (9, 10).
Therefore there has been a great amount of interest in early risk
stratification tools (11, 12) that can help identify those they may be
at risk of reccurence. Concurrently with this there has been a lot of
interest in new areas such as the gut bacterial microbiome for insight
into potential disease pathology.

There has been promising work on the bacterial microbiome and it's
ability to be able to complement existing screening methods such as
Fecal Immunoglobulin Test (FIT) or act alone as a screening tool (13,
14). There has also been research into how this microbiome could be
altered directly on tumor tissue itself (15). There have also been a few
studies that have shown how this microbiome (16) or specific members
within it (17) could be directly involved ith the pathogenesis of CRC.
These studies have helped to provide a tantilzing link between the
bacterial microbiome and CRC. However, at this present time there
remains limited information on the bacterial microbiome before and after
successful surgery for removal of the adenoma or carcinoma.

In this study we investigated what happened to the bacterial microbiome
before and after surgery for both adenoma and carcinoma individuals. Our
anlaysis includes both alpha and beta diversity analysis along with
investigating indvidual operational taxonomic units (OTUs). We next
utilized a Random Forest model that was trained on a completely separate
data set and observed how this model as well as specific OTUs within
this model performed on the initial and follow up samples. We then
created a new model to classify intial and follow up samples and
investigated how the model performs and what specific OUTs within the
model change. Finally, we observed how specific bacteria that have been
previously implicated in CRC changed before and after treatment in both
the adenoma and carcinoma individuals.

@@

\textbf{\emph{Bacterial Community and Fit Changes before and after
Treatment}} Based on thetayc distance metrics, comparing the initial to
the follow up samples, there was no difference between the adenoma and
carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There was a
difference in FIT between initial and follow up samples with the
carcinoma group having a significant decrease in FIT versus the adenoma
group (P-value = 2.15e-05) {[}Figure 1b{]}. Although the thetayc
distance metric change was similar between adenoma and carcinoma the
directionality of the change was significant in the cancer group between
initial and follow up (P-value = 0.002) but not for the adenoma group
(P-value = 0.997). This change can be visualized using an NMDS {[}Figure
1c \& 1d{]}. When all follow up samples were compared to each other
there was no significant overall difference between them (P-value =
0.085). There was no significant difference between initial and follow
up samples for observed OTUs, Shannon diversity, or evenness after
correction for multiple comparisons {[}Table S1{]}. Time of follow up
sample from initial sampling, did not have a significant difference
between adenoma and carcinoma (uncorrected P-value = 0.784).

\textbf{\emph{Outcome of Model Training}} The range of the AUC for model
training ranged from a minimum of 0.723 to a maximum of 0.795 with the
middle of all 100 runs having an AUC of 0.761. Interestingly, the worst
AUC model from training performed the best on it's respective 80/20
split test data {[}Figure 3{]}. In fact the 80/20 test performance
showed that the AUC for the middle model chosen was the most stable
(best training model test set AUC = 0.646, middle training model test
set AUC = 0.744, worse training model test set AUC = 0.904). That is to
say it had the smallest change in AUC in comparison to the minimum and
maximum AUC trained models. The middle model was close to the full
training data AUC NA. There was no significant difference between the
AUC of the best and middle training models (P-value = 1). There was also
no difference in the middle model versus worse (P-value = 0.0431) or
full data model (P-value = 1). The two comparisons with a significant
difference were between the worse training model and the best training
model (P-value = 6.83e-04) and the worse training model and full
training model (P-value = 1.2e-03).

\textbf{\emph{Most Important Variables to the Model}} Overall, there
were a total of 37 variables identified as being present in more than
50\% of the training models {[}Figure S2{]}. The top 5 most important
bacterial OTUs were Lachnospiraceae (Otu000013), Escherichia/Shigella
(Otu000018), Ruminococcaceae (Otu000020), Ruminococcus (Otu000017), and
Porphyromonas (Otu000153). These 5 OTUs were present in at least 90 out
of the total 100 different 80/20 runs.

\textbf{\emph{Surgical Removal of an Adenoma or Carcinoma Results in a
Decrease in Positive Probability Prediction}} A total of 1 sample was
omitted from the original 67 test sample set since it was missing a
complete set of follow up data. This left a total of 66 samples for test
predictions. After multiple comparison correction there was a
significant overall decrease in positive probability of a carcinoma and
adenoma (P-value = 1.11e-11) {[}Figure 4{]}. This decrease was
significant for both adenoma (P-value = 3.64e-11 {[}Figure 4a{]} and
carcinoma (P-value = 7.95e-08) {[}Figure 4b{]} alone. This decrease in
probability of lesion also held specifically for those with screen
relative neoplasias (SRN) (P-value = 7.63e-06). A total of 66 or 100\%
of all samples were correctly predicted to have a lesion. Although there
was a decrease in positive probability only 24 of the total 66
individuals were classified as adenoma or carcinoma free on follow up
(successful classification of 37.9\%). There was no significant
difference between the predictions and actual diagnosis for the initial
samples in the 67-sample cohort test set. However, the predictions were
significantly discordent with the diagnosis for the follow up samples
(P-value = 4.19e-10). Although there were discordent results the
respective sensitivity for the initial group was 100\% and for follow up
was 100\%, respectively.

There was 1 individual who still clearly had CRC on follow up as well as
5 individuals whose status on follow up was unknown. Although the 1
individual had a decrease in positive probability their follow up sample
was still higher than the cutoff threshold of 0.5 (positive probability
= 0.903). Interestingly, 1 individuals who were unknown on follow up
still were over the threshold cutoff of 0.5 even though, like the 1
individual with clear CRC on follow up, the probability of an adenoma or
carcinoma decreased {[}Table S2{]}.

The follow up positive probabilities were not affected by either
chemotherapy treatment (uncorrected P-value = 0.621) or radiation
therapy (uncorrected P-value = 0.255). There was also no difference in
the amount of change in the positive probability based on whether
individuals received chemotherapy (unccorected P-value = 0.718) or
radiation therapy (uncorrected P-value = 0.431).

\textbf{\emph{Specific OTUs in the Lesion Model are not Detected in
Follow Up Versus Initial Samples}} Overall, there were a total of 11
OTUs that were common between the main lesion model and the model for
classifying initial and follow up samples specifically {[}Table S3{]}. A
total of 1 OTU was still significant after multiple comparison
correction and it's lowest taxonomic identification was to Blautia. In
general, Otu000012 (Blautia) was decreased from initial to follow up
{[}Figure 5{]}. The relative abundance was not drastically different
then the mean of the values observed in the control training set
{[}Figure 5{]}.

\textbf{\emph{Differences in Adenoma and Carcinoma in Previously
Associated Cancer Bacteria}} First, there was a clear magnitude
difference in these specific OTUs based on whether they were from
adenoma or carcinoma individuals {[}Figure 6{]}. The carcinoma samples
showed a significant difference between initial and follow up samples
for Peptostreptococcus stomatis (P-value = 0.0183) and Porphyromonas
asaccharolytica (P-value = 0.0154) whereas there were no signficant
differences in any of these OTUs in the adenoma samples {[}Table S4{]}.

@@

We next created a model that incorporated both patient metadata, FIT,
and the bacterial microbiome to be able to predict lesions (adenoma or
carcinoma). Our middle training model, based on AUC, from 100 80/20
(train/test) splits was similar to the full training data model. It's
10-fold cross validated AUC was similar to it's test set AUC which was
not the case for both the best and worse training model {[}Figure 3{]}.
Using the full training data model we predicted the probability of a
lesion in the inital and follow up samples {[}Figure 4{]}. There was a
signficant decrease in positive probability regardless of whether the
sample was a carcinoma or adenoma. The overall sensitivity for lesion
detection in the intial samples was 100 and for follow ups was 100.
Although there was a decrease in overall probability of an adenoma or
carcinoma only 24 were below the 0.5 threshold out of the total 65
individuals who were diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in
our model {[}Figure 5 \& Table S3{]}. Many of the OTUs identified
classified to normal flora bacterium {[}Table S3{]}. Only a single OTU
though was significant after multiple comparison correction and the
lowest taxonomic identification of Otu000012 was to Blautia. Although
there was a difference in the relative abundance at initial and follow
up these values were not drastically different from the relative
abundance values observed in the control individuals of the training set
{[}Figure 5{]}. Although we were interested in what we could use to
classify those with either adenoma or carcinoma versus normal. We found
that the traditional bacteria associated with CRC were higher in
magnitude in the carcinoma group and there were significant differences
in some of these OTUs between the initial and follow up samples
{[}Figure 5 and Tabl S5{]}. This research provides evidence that it is
possible to use bacterial microbiome data to create a highily sensitive
model, that is reactive to therapy, for detection of adenoma or
carcinoma. It accomplishes this by using a unique sample set in which
before and after surgery stool samples are available for assessment. By
using these types of samples we are not only able to show sensitivity of
lesion prediction but also able to show that this model is reactive.
That is to say that after surgery for removal of the adenoma or
carcinoma it decreases the positive probability to reflect a lower
likelihood of the individual having an adenoma or carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al (13).
In brief, study exclusion involved those who had already undergone
surgery, radiation, or chemotherapy, had colorectal cancer before a
baseline stool sample could be obtained, had IBD, a known hereditary
non-polyposis colorectal cancer, or Familial adenomatous polyposis.
Samples used to build the model used for prediction were collected
either prior to a colonoscopy or between 1 - 2 weeks after. The
bacterial microbiome has been shown to nomralize within this time period
(18). Kept apart from this training set were a total of 67 individuals
that not only had a sample as described previoulsy but also a follow up
sample between 188 - 546 days after surgery and treatment had been
completed. This study was approved by the University of Michigan
Institutional Review Board. All study participants provided informed
consent and the study itself conformed to the guidelines set out by the
Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) (19). 16S rRNA gene
sequencing was completed as previously described by Kozich, et al. (20).
In brief, DNA extraction used the 96 well Soil DNA isolation kit (MO BIO
Laboratories) and an epMotion 5075 automated pipetting system
(Eppendorf). The V4 variable region was amplified and the resulting
product was split between three sequencing runs with control, adenoma,
and carcinoma evenly represented on each run. Each group was randomly
assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described (20). The general processing workflow
using mothur is as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screening for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The numer of sequences for each sample
was rarified to 10521 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest (21) algorithm
was used to create the model used for prediction of lesion (adenoma or
carcinoma) for the 67 individuals with follow up samples. The model
included data on FIT and the bacterial microbiome. Non-binary data was
checked for near zero variance and auto correlation. Data columns that
had near zero variance were removed. Columns that were correlated with
each other over a Spearman correlation coefficient of 0.75 had one of
the two columns removed. This pre-processing was performed with the R
package caret (v6.0.73). Optimization of the mtry hyperparameter
involved taking the samples and making 100 80/20 (train/test) splits in
the data where control and lesion were equally represented in the 80 and
20 split, respectively. This 80\% portion was then split again into an
80/20 split, and run through 20 repeated 10-fold cross validations to
optimzie the model's AUC (Area Under the Curve of the Receiver Operator
Characteristic). This resulting model was then tested on the 20\% of the
data that was originally held out from this overall process. Once the
ideal mtry was found the entire 490 sample set was used to create the
final Random Forest model on which testing on the 67-person cohort was
completed. The default cutoff of 0.5 was used as the threshold to
classify individuals as positive or negative for lesion. The
hyperparameter, mtry, defines the number of variables to investigate at
each split before a new division of the data is created.

@@

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA (22) in the vegan package
(v2.4.1) while comparisons between ROC curves utilized the method by
DeLong et al. (23) executed by the pROC (v1.8) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied (24) and a corrected P-value of less than 0.05
was considered significant. Unless otherwise stated the P-values
reported are those of the BH corrected ones.

@@

\subsection{Acknowledgements}\label{acknowledgements}

The authors thank the Great Lakes-New England Early Detection Research
Network for providing the fecal samples that were used in this study.
This study was supported by funding from the National Institutes of
Health to P. Schloss (R01GM099514, P30DK034933) and to the Early
Detection Research Network (U01CA86400).

@@

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Initial and Follow Up}* A) Positive probability adjustment
of those with carcinoma from intial to follow up sample B) Positive
probability adjustment of those with adenoma as well as those with SRN
and the probability adjustment from initial to follow up sample. The
dotted line represents the threshold used to make the decision of
whether a sample was lesion positive or not.

\textbf{Figure 5: Lesion Model OTU with a Significant Decrease in
Relative Abundance that is also Predictive of Initial and Follow Up.}
After multiple comparison correction 1 (Blautia) was the only one with a
P-value \textless{} 0.05. The dotted line represents the average
relative abundance in the control training group.

@@

\textbf{Figure S2: Summary of Important Variables in both Lesion and
Initial/Follow Up models} A) MDA of the most important variables in the
lesion model. The black point represents the median and the different
colors are the different runs up to 100. B) The total number of
appearances of each variable in the 100 different lesion models. The
cutoff of 50\% was used to assess importance.

\textbf{Figure S3: Thetayc Graphed Against Time of Follow up Sample from
Initial}

>>>>

\textbackslash{}begin\{center\}\\
\vspace{25mm}

Running Title: Human Microbiome and Colorectal Cancer

@@

Marc A Sze\textsuperscript{1}, Nielson T Baxter\textsuperscript{2}, Mack
T Ruffin IV\textsuperscript{3}, Mary AM Rogers\textsuperscript{4}, and
Patrick D Schloss\textsuperscript{1}\({^\dagger}\)

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

\textsuperscript{1}Department of Microbiology and Immunology, University
of Michigan, Ann Arbor, MI\\
\textsuperscript{2}Department of Internal Medicine, Division of
Infectious Disease, University of Michigan, Ann Arbor, MI\\
\textsuperscript{3}Department of Family Medicine and Community Medicine,
Penn State Hershey Medical Center, Hershey, PA\\
\textsuperscript{4}Department of Internal Medicine, University of
Michigan, Ann Arbor, MI \textbackslash{}end\{center\}

@@

\textbf{Background:} Colorectal cancer (CRC) continues to be a worldwide
health problem with early detection being used as a key component in
mitigating deaths due to the disease. Previous research suggests a link
between stool bacterial microbiome and CRC. The overall objective was to
investigate the changes in the bacterial microbiome after surgery in
patients with lesion (i.e.~adenoma or carcinoma). Specifically, we
wanted to identify what within the community was different within those
undergoing surgical removal of lesion. We also investigated the use of
the bacterial microbiome and Fecal Immunoglobulin Test (FIT) to build
models to either classify individuals as having a lesion or whether,
based on the bacterial microbiome, the sample could be classified
correctly as before or after surgery.\\
\textbf{Results:} Adenoma individual's bacterial microbiome were more
similar to their pre-surgery sample then those with carcinoma (P-value =
0.00198) and this was also reflected in FIT as well (P-value =
2.15e-05). There was no significant difference in any indivdiual OTU
between samples before and after surgery (P-value \textgreater{} 0.125.
A model with a total of 37 variables was able to classify lesion with an
AUC range of 0.847 to 0.791 while the model to classify samples as
before and after had 33 with an AUC range of 0.79 to 0.651 for 100 20
repeated 10-fold cross-validated runs. Both models had a significant
decrease in the positive probability of a lesion between individual's
before versus after surgery samples (P-value = 1.91e-11 and 6.72e-12).
In total there were 14 OTUs that were common to both models and were
mostly commensals with largest representation from OTUs belonging to
Bacteroides, Blautia, Streptococcus, and Clostridiales.\\
\textbf{Conclusions:} Our data suggests that treatment not only
significantly reduces the probability of having a colonic lesion but
also causes detectable changes in the bacterial microbiome. Further
surveillance of these individuals will enable us to determine whether
models such as the one we present here can also be used to predict
recurrence of colorectal cancer.

@@

\subsubsection{Keywords}\label{keywords}

bacterial microbiome; colorectal cancer; polyps; FIT; detection; risk
factors

@@

\subsection{Background}\label{background}

Colorectal cancer (CRC) continues to be a leading cause of cancer
related deaths and is the second most common cancer death among men aged
40-79 years of age {[}1,2{]}. Over the last few years death due to the
disease has seen a significant decrease, thanks mainly to improvements
in screening {[}1{]}. However, despite this improvement there are still
approximately 50,000 deaths from the disease a year {[}2{]}. It is
estimated that around 5-10\% of all CRCs can be explained by autosomal
dominant inheritance {[}3{]}. The vast majority of CRCs are not
inherited and the exact etiology of the disease has not been well worked
out {[}2{]}. Although many risk factors have been identified {[}2{]} and
non-invasive screening techniques have started to be put into consistent
use {[}4,5{]} there has been an consistent additional increase in the
incidence of CRC in the younger population.

This increased incidence of CRC in the younger population is concerning
since having either an adenoma or carcinoma increases ones risk for
future adenomas or carcinomas {[}6--8{]}. This increased risk can also
carry with it an increased chance of mortality due to this recurrence
{[}9,10{]}. Therefore there has been a great amount of interest in early
risk stratification tools {[}11,12{]} that can help identify those they
may be at most susceptibility to reccurence. Concurrently, there has
also been a lot of interest in new areas that could have a role in
disease pathogenesis, such as the gut bacterial microbiome.

There has been promising work on the bacterial microbiome and it's
ability to be able to complement existing screening methods such as
Fecal Immunoglobulin Test (FIT) or act alone as a screening tool
{[}13,14{]}. There has also been research into how this microbiome could
be altered directly on tumor tissue itself {[}15{]}. A few studies have
now even shown how this microbiome {[}16{]} or specific members within
it {[}17{]} could be directly involved with the pathogenesis of CRC.
These studies have helped to provide a tantilzing link between the
bacterial microbiome and CRC. However, at this present time there
remains limited information on the bacterial microbiome before and after
successful surgery for removal of the adenoma or carcinoma and whether
it changes at all.

In this study we investigated what happened to the bacterial microbiome
before and after surgery for both adenoma and carcinoma individuals. Our
anlaysis includes both alpha and beta diversity analysis along with
investigation of individual operational taxonomic units (OTUs). We also
utilized Random Forest models and observed how these models well as
specific OTUs within this model performed on before (initial) and after
(follow up) surgery samples. We used these models to look for similar
important OTUs to identify the most important ones for not only
classifying initial and follow up samples but also lesion or normal.

@@

\textbf{\emph{Bacterial Community and Fit Changes before and after
Treatment}} Based on thetayc distance metrics, comparing the initial to
the follow up samples, there was a significant difference between the
adenoma and carcinoma groups (P-value = 0.00198) {[}Figure 1a{]}. There
was also a significant difference in FIT between initial and follow up
samples with the carcinoma group having a significant decrease in FIT
versus the adenoma group (P-value = 2.15e-05) {[}Figure 1b{]}. The whole
community structure before and after surgery are visualized on NMDS
graphs for both adenoma {[}Figure1c{]} (PERMANOVA = 0.002) and carcinoma
{[}Figure 1d{]} (PERMANOVA = 0.997). When all initial and follow up
samples were compared to each other there was no significant overall
difference between them (PERMANOVA = 0.085). There was also no
significant difference between initial and follow up samples for
observed OTUs, Shannon diversity, or evenness after correction for
multiple comparisons {[}Table S1{]}. There was no significant difference
between initial and follow up samples for any single OTU {[}Figure
S1{]}.

\textbf{\emph{Differences in Adenoma and Carcinoma of Previously
Associated Cancer Bacteria}} We next examined whether there were
differences in previously well described carcinoma associated OTUs.
These included the OTUs that aligned with Porphyromonas asaccharolytica
(Otu000153), Fusobacterium nucleatum (Otu000226), Parvimonas micra
(Otu000460), and Peptostreptococcus stomatis (Otu000653). First, the
carcinoma samples showed a significant difference between initial and
follow up samples for Peptostreptococcus stomatis (P-value = 0.0183) and
Porphyromonas asaccharolytica (P-value = 0.0154) whereas there were no
signficant differences in any of these OTUs in the adenoma samples
{[}Table S4{]}. Second, when these OTUs were present, there was a clear
magnitude difference in these specific OTUs based on whether they were
from adenoma or carcinoma individuals {[}Figure 2{]}. However, only a
small percentage of those with adenoma or carcinoma were positive for
any of these OTUs.

\textbf{\emph{Full and Reduced Model Results}} Since differences were
observed between initial and follow up samples and only a small number
of individuals were positive for previously associated CRC bacteria we
next investigated if we could create a model that could adequately
classify and adjust lesion probability based on the bacterial community
and FIT. The lesion model had an AUC range of 0.723 to 0.795 versus the
initial follow up model which had and AUC range of 0.451 to 0.67 after
100 20 repeated 10-fold cross validations. Interestingly, identification
of the most important variables and reducing the variables considered to
only these increased the AUC in the lesion model (0.791 - 0.847) and
initial and follow up model (0.651 - 0.79).

The test set AUC range for the full and reduced lesion model were
similar to that reported for the training set AUC ranges and the ROC
curve ranges overlap each other {[}Figure 3a{]}. The ROC curve for the
final lesion model used falls within the range of both the full and
reduced lesion model {[}Figure 3a{]}. Interestingly, the test set AUC
range for the initial and follow up performed much better then the
training set AUCs. Both the full and reduced initial and follow up
models overlapped with each other {[}Figure 3b{]} there was a marked
decrease in the ROC curve for the final model used.

\textbf{\emph{Most Important Variables to the Models}} The reduced
models were built based on the most important variables to the
respective full model. For the lesion model there were a total of 37
variables {[}Figure S2{]} whereas for the initial and follow up model
there were a total of 33 variables {[}Figure S3{]}. For both models FIT
resulted in the largest decrease in MDA {[}Figure S2a \& S3a{]}.

\textbf{\emph{Positive Probability Prediction after Surgical Removal of
Adenoma or Carcinoma}} Regardless of model used there was a significant
decrease in the positive probability of either the sample being lesion
or an initial sample on follow up {[}Figure 4 \& S4{]} (full lesion
P-value = 1.11e-11, reduced lesion P-value = 1.91e-11, initial and
follow up P-value = 6.71e-12, and reduced initial and follow up P-value
= 6.72e-12).

For the full and reduced lesion model there was a significant difference
in the classification for the lesion model between predicted and actual
(P-value = 4.19e-10 and 6.98e-10, respectively) but not for the initial
follow up model (P-value = 1.00 and 1.00). However, the lesion model
correctly kept the one individual who still had a carcinoma on follow up
above the cut off point {[}Figure 4a \& S4a{]} for a positive call while
the initial and follow up models did not {[}Figure 4b \& S4b{]}.

\textbf{\emph{Common OTUs to both Lesion and Initial and Follow Up
Models}} There were a total of 14 OTUs that were common to both models.
Of these OTUs the most common taxonomic identifications were to Blautia,
Bacteroides, Streptococcus, and Clostridiales. The majority of these
OTUs had taxonomic identification to bacteria typically thought of as
commensal {[}Table S2{]}.

\textbf{Treatment and Time Differences}

There was no difference in the amount of change in positive probability
for either the full or reduced lesion model for either chemotherapy
(P-value = 0.821 and 0.821) or radiation therapy (P-value = 0.69 and
0.981). Although the initial follow up model was similar there was a
significant decrease in positive probability for those treated with
chemotherapy (P-value = 7.04e-04 and 5.07e-03Time of follow up sample
from initial sampling, did not have a significant difference between
adenoma and carcinoma (uncorrected P-value = 0.784).

@@

We next created a model that incorporated both patient metadata, FIT,
and the bacterial microbiome to be able to predict lesions (adenoma or
carcinoma). Our middle training model, based on AUC, from 100 80/20
(train/test) splits was similar to the full training data model. It's
10-fold cross validated AUC was similar to it's test set AUC which was
not the case for both the best and worse training model {[}Figure 3{]}.
Using the full training data model we predicted the probability of a
lesion in the inital and follow up samples {[}Figure 4{]}. There was a
signficant decrease in positive probability regardless of whether the
sample was a carcinoma or adenoma. The overall sensitivity for lesion
detection in the intial samples was and for follow ups was . Although
there was a decrease in overall probability of an adenoma or carcinoma
only were below the 0.5 threshold out of the total individuals who were
diagnosed as not having a carcinoma on follow up.

We then investigated which OTUs could potentially be more important in
our model {[}Figure 5 \& Table S3{]}. Many of the OTUs identified
classified to normal flora bacterium {[}Table S3{]}. Only a single OTU
though was significant after multiple comparison correction and the
lowest taxonomic identification of was to . Although there was a
difference in the relative abundance at initial and follow up these
values were not drastically different from the relative abundance values
observed in the control individuals of the training set {[}Figure 5{]}.
Although we were interested in what we could use to classify those with
either adenoma or carcinoma versus normal. We found that the traditional
bacteria associated with CRC were higher in magnitude in the carcinoma
group and there were significant differences in some of these OTUs
between the initial and follow up samples {[}Figure 5 and Tabl S5{]}.
This research provides evidence that it is possible to use bacterial
microbiome data to create a highily sensitive model, that is reactive to
therapy, for detection of adenoma or carcinoma. It accomplishes this by
using a unique sample set in which before and after surgery stool
samples are available for assessment. By using these types of samples we
are not only able to show sensitivity of lesion prediction but also able
to show that this model is reactive. That is to say that after surgery
for removal of the adenoma or carcinoma it decreases the positive
probability to reflect a lower likelihood of the individual having an
adenoma or carcinoma.

@@

\textbf{\emph{Study Design and Patient Sampling}} The sampling and
design of the study was similar to that reported in Baxter, et al
{[}13{]}. In brief, study exclusion involved those who had already
undergone surgery, radiation, or chemotherapy, had colorectal cancer
before a baseline stool sample could be obtained, had IBD, a known
hereditary non-polyposis colorectal cancer, or Familial adenomatous
polyposis. Samples used to build the model used for prediction were
collected either prior to a colonoscopy or between 1 - 2 weeks after.
The bacterial microbiome has been shown to nomralize within this time
period {[}18{]}. Kept apart from this training set were a total of 67
individuals that not only had a sample as described previoulsy but also
a follow up sample between 188 - 546 days after surgery and treatment
had been completed. This study was approved by the University of
Michigan Institutional Review Board. All study participants provided
informed consent and the study itself conformed to the guidelines set
out by the Helsinki Declaration.

\textbf{\emph{Fecal Immunochemical Test and 16S rRNA Gene Sequencing}}
FIT was analyzed as previously published using both OC FIT-CHEK and
OC-Auto Micro 80 automated system (Polymedco Inc.) {[}19{]}. 16S rRNA
gene sequencing was completed as previously described by Kozich, et al.
{[}20{]}. In brief, DNA extraction used the 96 well Soil DNA isolation
kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting
system (Eppendorf). The V4 variable region was amplified and the
resulting product was split between three sequencing runs with control,
adenoma, and carcinoma evenly represented on each run. Each group was
randomly assigned to avoid biases based on sample collection location.

\textbf{\emph{Sequence Processing}} The mothur software package
(v1.37.5) was used to process the 16S rRNA gene sequences. This process
has been previously described {[}20{]}. The general processing workflow
using mothur is as follows: Paired-end reads were first merged into
contigs, quality filtered, aligned to the SILVA database, screening for
chimeras, classified with a naive Bayesian classifier using the
Ribosomal Database Project (RDP), and clustered into Operational
Taxonomic Units (OTUs) using a 97\% similarity cutoff with an average
neighbor clustering algorithm. The numer of sequences for each sample
was rarified to 10521 in an attempt to minimize uneven sampling.

\textbf{\emph{Lesion Model Creation}} The Random Forest {[}21{]}
algorithm was used to create the model used for prediction of lesion
(adenoma or carcinoma) for the 67 individuals with follow up samples.
The model included data on FIT and the bacterial microbiome. Non-binary
data was checked for near zero variance and auto correlation. Data
columns that had near zero variance were removed. Columns that were
correlated with each other over a Spearman correlation coefficient of
0.75 had one of the two columns removed. This pre-processing was
performed with the R package caret (v6.0.73). Optimization of the mtry
hyperparameter involved taking the samples and making 100 80/20
(train/test) splits in the data where control and lesion were equally
represented in the 80 and 20 split, respectively. This 80\% portion was
then split again into an 80/20 split, and run through 20 repeated
10-fold cross validations to optimzie the model's AUC (Area Under the
Curve of the Receiver Operator Characteristic). This resulting model was
then tested on the 20\% of the data that was originally held out from
this overall process. Once the ideal mtry was found the entire 490
sample set was used to create the final Random Forest model on which
testing on the 67-person cohort was completed. The default cutoff of 0.5
was used as the threshold to classify individuals as positive or
negative for lesion. The hyperparameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
is created.

@@

\textbf{\emph{Statistical Analysis}} The R software package (v3.3.2) was
used for all statisitical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}22{]} in the vegan package
(v2.4.1) while comparisons between ROC curves utilized the method by
DeLong et al. {[}23{]} executed by the pROC (v1.8) package. Comparisons
between probabilities as well as overall amount of OTU between initial
and follow up samples utilized a paired wilcoxson ranked sum test. Where
multiple comparison testing was needed a Benjamini-Hochberg (BH)
correction was applied {[}24{]} and a corrected P-value of less than
0.05 was considered significant. Unless otherwise stated the P-values
reported are those of the BH corrected ones.

@@

@@

\textbf{Figure 4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial and Follow Up for Reduced Variable
Models} A) Positive probability adjustment of those with carcinoma from
intial to follow up sample B) Positive probability adjustment of those
with adenoma as well as those with SRN and the probability adjustment
from initial to follow up sample. The dotted line represents the
threshold used to make the decision of whether a sample was lesion
positive or not.

@@

\textbf{Figure S2: Summary of Important Variables in the Lesion Model}
A) MDA of the most important variables in the lesion model. The black
point represents the median and the different colors are the different
runs up to 100. B) The total number of appearances of each variable in
the 100 different lesion models. The cutoff of 50\% was used to assess
importance.

\textbf{Figure S3: Summary of Important Variables in Initial Follow Up
Model} A) MDA of the most important variables in the lesion model. The
black point represents the median and the different colors are the
different runs up to 100. B) The total number of appearances of each
variable in the 100 different lesion models. The cutoff of 50\% was used
to assess importance.

\textbf{Figure S4: Breakdown by Carcinoma and Adenoma of Prediction
Results for Lesion and Initial and Follow Up for Full Variable Model} A)
Positive probability adjustment of those with carcinoma from intial to
follow up sample B) Positive probability adjustment of those with
adenoma as well as those with SRN and the probability adjustment from
initial to follow up sample. The dotted line represents the threshold
used to make the decision of whether a sample was lesion positive or
not.

\textbf{Figure S5: Thetayc Graphed Against Time of Follow up Sample from
Initial}

\subsection{Declarations}\label{declarations}

\subsubsection{Ethics approval and consent to
participate}\label{ethics-approval-and-consent-to-participate}

\subsubsection{Consent for publication}\label{consent-for-publication}

\subsubsection{Availability of data and
material}\label{availability-of-data-and-material}

\subsubsection{Competing Interests}\label{competing-interests}

All authors declare that they do not have any relevent competing
interests to report.

\subsubsection{Funding}\label{funding}

This study was supported by funding from the National Institutes of
Health to P. Schloss (R01GM099514, P30DK034933) and to the Early
Detection Research Network (U01CA86400).

\subsubsection{Authors' contributions}\label{authors-contributions}
